Renewal and stability in populations structured by remaining years of life

### 2014-09-11 ThanoRepro/ThanoRepro.tex
changes throughout text

A lifetable approach to transforming chronologically classified data into
thanatologically classified data is valid when the given lifetable accurately
describes the future mortality of age-classified population. To make
a projective statement about a current stock of population, one does well to
factor in future changes in mortality, and to allow the trajectory of changes
to vary between ages/cohorts. In the case of a stable population model,
mortality is assumed invariant over time, which means that the same survival
pattern underlies each chronological age class, and we therefore draw from a
single lifetable. Under these conditions, data classified by chronological age
can be decomposed and redistributed into thanatological age classes with a
single mortality pattern. The basic operation is to multiply the probability of
surviving from exact chronological age $a$ to exact age $a+y$,
$\frac{l(a+y)}{l(a)}$, and then dying at exact age $a+y$, $\mu(a+y)$. The
index $y$ in this case means $y$ years in the future. For age 0 this probability
is simply the $d(a)$ column of the lifetable, standardized to sum to 1. For higher ages, simply condition on survival:\footnote{In the reliability literature, our concept of $d(y|a)$ is denoted by \begin{equation} X - t | X > t
\end{equation}
where $X$ is the total lifespan and $t$ is the current chronological age
attained.}

@@

\begin{equation}
d_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \quad\quad \text{,}
\end{equation}

\noindent where $\omega$ is the highest age. One
produces in this way a probability density function over future death times
for each chronological age, $a$, and the $d(y|a)$- weighted average of $y$ for
a given $a$ returns the familiar remaining life expectancy column of the lifetable,
$e(a)$. If $P(a)$ is an chronological age-classified population count,
we can derive $P(y)$, a thanatological age-classified population count, as
follows:

@@

Since $1 = \int_0^\infty d(y|a) \dd y$ for each $a$,
this density is useful for decomposing data classified by chronological age. For
example, we can break down a population pyramid into thanatological age classes.
Figure~\ref{fig:USdecomp} shows this exercise for 2010 US data, assuming the
2010 period mortality schedule remains constant. When looking from the
chronological age perspective, this decomposition reveals thanatological
heterogeneity, but when viewed from the thanatological perspective
(Figure~\ref{fig:USdecomp}) it reveals chronological age heterogeneity and the
overall thanatological age profile, the future death flow of the current
population.
Figure~\ref{fig:USrecomp} shows this \textit{re}composition for the same 2010
US data.

@@

It bears repeating that lifetable-derived thanatological age structure is probabilistic until after death, since it refers to the future according to an assumed mortality pattern.
As such, Figure~\ref{fig:USrecomp} is projective in nature, a forward-looking
glance at population attrition given the population stock and lifetable of a
particular moment, whereas Figure~\ref{fig:USdecomp} is reflective in nature, since a population's chronological age-structure is mostly the fruit of past fertility, but also migration and to a lesser degree attrition. Note that any count classified by chronological age can be
reclassified in this way, as long as you have a lifetable that plausibly represents the population whose data you wish to reclassify. This means we can derive thanatological fertility rates, $F_y$, by applying \eqref{eq:transform}
to age-classified birth counts, $B_a$, to get $B_y$ and again to
exposure-to-risk, $E_a$ (here we take exposures from all ages), to get $E_y$ and then dividing:

\begin{equation}
f^\star_y = \frac{B_y}{E_y}
\end{equation}

We are interested to use thanatological fertility rates, $F_y$ in
models of population renewal,\footnote{This is probably premature, but helps to
understand the nature and provenance of model components.} rather than to
explicitly study the nature of these rates. Nonetheless it is best not to throw such rates into a model blindly, so we offer a schematic overview of some of their characteristics.
Figure~\ref{fig:FySpaghetti} represents the full variety in
female thanatological period fertility rates that can be found for all years
of data that overlap in the HFD and HMD.\footnote{There are as of this writing
1834 population-years of overlap between the HMD and HFD, including a wide
variety of fertility and mortality combinations.} Figure~\ref{fig:FxSpaghetti} gives ASFR for the same
populations and years as a more familiar reference, but note that both scales
are different! With no indication of particular populations or time series, one
already concludes that period thanatological fertility rates have a
characteristic shape and are not random, erratic or informationless: There is a
pattern to fertility rates by remaining years of life, and it varies over time
and between populations within some range of normality. Thanatological fertility
rates have a wider distribution than do chronological rates. Note that this
spaghetti plot includes both fertility booms and busts, as well as some
mortality crises (1918, WWII). Some patterns to note, but which we do not
separate graphically here, are that the left tail (a gauge of how orphan-prone a
population is) has tended to fall over time. All populations have shown a
rightward shift in both the mean and modal mother's thanatological age at birth
(TAB) over time (that's unambiguously a good thing). Several populations now
show modal TABs of over 60 years. Thanatological total fertility rates (TTFR)
track standard period total fertility rates (PTFR) rather closely, but tend to
be somewhat higher (not shown).

@@

In general, such fertility rates will not be palatable for purposes of
projection, unless the demographer believes that their empirical regularity is
somehow stronger than typical ASFR --- an argument we do not make. There may be
alternative ways of defining thanatological fertility rates, for instance by
transforming from a stable population, which would make the resulting model more
conformable with the Lotka renewal model. First let us define the basic model.

\FloatBarrier
\section*{The thanatological renewal model}
For the renewal model that follows, we are most interested in using the kind
of fertility rates shown in Figure~\ref{fig:FySpaghetti}, and we refer to a
unisex population based on female vital rates. In the following, I
explain describe the model in four different ways: as a continuous
model, as a matrix model, with a lifecylcle graph, and intuitively by
appealing to how population pyramids and leaves (described earlier) change over
time. Assuming constant vital rates, the births for the present year are given by
\begin{align}
B(t) =& \int _0^\infty P(y,t)f^\star(y) \dd y = \int _0^\infty P(a,t)f(a) \dd a
\quad\text{,} \intertext{where $a$ indexes chronological age, $y$ indexes
thanatological age, $P(a), P(y)$ are population counts, and $f^\star(y), f(a)$ are
exact specific fertility probabilities (rates), which here are best thought of
as the fertility rates of females born to females. The thanatological integral can
be broken down back in terms of chronological age:} =& \int_{y=0}^\infty
\int_{a=0}^\infty f^\star(y)\frac{P(a,t)d(a+y)}{l(a)} \dd a \dd y \quad\text{,}
\intertext{where $d(a)$ is the continuous lifetable death distribution with
radix of 1, so $d(a)/l(0)$. We can relate the present population to past births
with $P(a,t) = B(t-a)l(a)$:}
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t-a)d(a+y)\dd a \dd y
\quad\text{.} \intertext{Eventually-- usually a bit quicker than is the case for
chronological age-- strong ergodicity will assert itself, and $B(t)$ will be related to
$B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth:}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\quad\text{.}
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation:}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) d(a+y)e^{-ra}\dd a \dd y
\quad\text{.}
\end{align}
Now compare this to Lotka's chronological formulation
\begin{align}
\label{eq:lotka}
1 =& \int_{a=0}^\infty f(a)l(a)e^{-ra}\dd a \quad\text{,}
\intertext{which can be expressed as well as:}
  =& \int_{a=0}^\infty \int_{y=0}^\infty f(a)d(a+y)e^{-ra}\dd y \dd a
\end{align}
and note that these are really quite similar, since $\int _{y=0}^\infty
d(a+y)\dd a = l(a)$. For intuition, notice that $l(a)$ is here split
up into pieces of $d(a)$, and imagine a 2D surface of these, where one axis is chronological age and the
other axis is thanatological age. For the chronological case \eqref{eq:lotka},
we multiply chronological age-specific fertility rates, $f(a)$ over one margin,
and for the thanatological case we multiply $f^\star(y)$ over the other margin.
There is a strong parallel here with the case of two-sex age-specific fertility rates. As in the case of
divergence between male and female single-sex models, there will always be divergence between
single-sex models under chronological versus thanatological age, unless the
starting population is stable. A proof of equivalence in the case of
$f^\star(y)$ being calculated from a stable starting population is provided
in Appendix~\ref{app:B}. Even though the modeled population stocks are in a way
commensurable, the rates used here, $f(a)$ versus $f^\star(y)$ are calculated on the basis of differently distributed denominators $E(a)$ versus $E(y)$.

There are gaps in the above line of development, since the jump from
\eqref{eq:ergostep1} to \eqref{eq:ergostep2} (strong ergodicity) is unproven,
although it is rather intuitive, given the much greater density of connections
within the model; persons from nearly any thanatologcal age\footnote{I say
\textit{nearly} because the very highest thanatological ages tend to be
composed of pre-menarchical females, so these can be thought of as structural
zeros. I am uncertain if it is a record, but it is certainly an extreme case
useful for illustration: Jeanne Calment, who lived to 122, had a daughter at age
22, which means she was about thanatological age 100
(\url{http://en.wikipedia.org/wiki/Jeanne_Calment}). If the lifetable used to
transform to thanatological closes out at age 110+, as is the case with HMD data, then fertility rates for thantological ages 101-110+ would be structural
zeros.} can produce offspring that can have any other thanatological age. In
this sense, the smoothing mechanism at play as the population passes through
time must be much stronger than that for chronological age, at least in most
cases. A further conjecture is that the jump will also hold for the case of weak ergodicity. The mechanisms at play unfold in the same way as those so intuitively described by \citet{arthur1982ergodic}, and said proof may apply here without further modification. A proof of the uniqueness of the solution to \eqref{eq:thanoren} is given in Appendix~\ref{app:A}.

@@

As with Leslie matrices, the above projection matrix may be manipulated using
standard matrix techniques. Where $\textbf{p}$ is our population vector, we
project by multiplying $\textbf{Y}$ from the left: $\textbf{p}(t + 1) =
\textbf{Y}\textbf{p}(t)$. From $\textbf{Y}$, we can extract such information as 
the intrinsic growth rate, $r$ (natural log of the largest
real eigenvalue), the stable thanatological age-structure (the real part of the
eigenvector that corresponds to the largest real eigenvalue), or the pace of
convergence to stability (the ratio of the \nth{1} to the \nth{2} eigenvalues).
\footnote{See \citet[p.86-87]{caswell2001matrix}.}

The
thanatological lifecycle graph is directed and strongly connected. The matrix is
also redudcible because each thanatological age class is reachable from every
other thanatological age class, either directly via fertility, or else by ageing toward lower ages. The thanatological matrix is also
easy to identify as primitive, since it contains self-loops (fertility elements
on the matrix diagonal). The matrix is said to be regular, since we can
imagine some value $k$ where each thanatological age class is $k$ steps away
from every other thanatological age class. In the fertility block of the matrix
(all rows, but not necessarily all columns), $k$ is 0, due to direct connections
between all fertile thanatological ages, but in order to make this condition hold for pre-menarcical females with very long lives, $k$ must be equal to the maximum lifespan. This can also be demonstrated simply by
setting $\textbf{Y}^k$ to some large value such that all elements of $\textbf{Y}
> 0$.

@@

\section{An iterative method to find the thanatological $r$}
\citet{coale1957new} proposed a fast-converging iterative approach to estimate
the intrinsic growth rate for the Lotka-Euler equation. For the thanatological
renewal model, a similar approach may be taken, with some slight
modifications to Coale's original.
The following steps can be followed to estimate $r$ from
Equation~\eqref{eq:thanoren}:

>>>>

A lifetable approach to transforming chronologically classified data into
thanatologically classified data is valid when the given lifetable accurately
describes the future mortality of age-classified population. To make
a projective statement about a current stock of population, one does well to
factor in future changes in mortality, and to allow the trajectory of changes
to vary between ages/cohorts. In the case of a stable population model,
mortality is assumed invariant over time, which means that the same survival
pattern underlies each birth cohort. This is the case we explore, and so the
following formulas assume a single lifetable.

The basic transformation to thanatological age proceeds by defining a
deaths density function, similar the lifetable $d(a)$, but over two
dimensions: chronological age attained, $a$, and remaining years of life, $y$.
This function we denote $d(y|a)$ --- The density distribution of thanatological
age depends on chronological age. I use the same function name $d()$, because
it is essentially the same notion as the familiar $d(a)$, but rescaled as a
function of chronological age. For ages higher than 0, simply condition
on survival:\footnote{In the reliability literature, our concept of $d(y|a)$ is denoted by \begin{equation} X - t | X > t
\end{equation}
where $X$ is the total lifespan and $t$ is the current chronological age
attained.}

@@

\begin{equation}
\label{eq:dyadiscret}
d_{y, a} =  \frac{d_{a+y}+d_{a+y+1}}{L_a + L_{a+1}} \quad\quad \text{,}
\end{equation}

\noindent where $\omega$ is the highest
age.\footnote{Equation~\eqref{eq:dyadiscret} ill incur some error depending on
the method used to derive $L_a$. To remove this error, rescale over the $a$ margin so that each distribution conditioned on $a$
sums to 1. Working with $l_a$ rather than $L_a$ eliminates error and yields
similar results, but is not necessarily better.} One produces in this way a
probability density function over future death times for each chronological age,
$a$. The $d(y|a)$- weighted average of $y$ for a given $a$ returns the familiar
remaining life expectancy column of the lifetable, $e(a)$. If $P(a)$ is a
chronological age-classified population count, we can derive $P(y)$, a
thanatological age-classified population count using ~\eqref{eq:transform}.

@@

Since $1 = \int_0^\infty d(y|a) \dd y$ for each $a$,
this density is useful for decomposing data classified by chronological age. For
example, we can break down a population pyramid into thanatological age classes.
Figure~\ref{fig:USdecomp} shows this exercise for 2010 US data, assuming the
2010 period mortality schedule remains constant. When looking from the
chronological age perspective, this decomposition reveals thanatological
heterogeneity, but when viewed from the thanatological perspective
(Figure~\ref{fig:USdecomp}) it reveals chronological age heterogeneity and the
overall thanatological age profile, the future death flow of the current
population.
Figure~\ref{fig:USrecomp} shows this recomposition for the same 2010
US data.

@@

It bears repeating that lifetable-derived thanatological age structure is probabilistic until after death, since it refers to the future according to an assumed mortality pattern.
As such, Figure~\ref{fig:USrecomp} is projective in nature, a forward-looking
glance at population attrition given the population stock and lifetable of a
particular moment, whereas Figure~\ref{fig:USdecomp} is reflective in nature,
since a population's chronological age-structure is mostly the fruit of past
fertility, but also migration and to a lesser degree attrition. Any count
classified by chronological age can be reclassified in this way given a
corresponding lifetable. Exploiting this, we derive thanatological fertility
rates, $f^\star(y)$, by applying \eqref{eq:transform} to age-classified birth
counts, $B(a)$, to get $B(y)$ and again to exposure-to-risk, $E(a)$ (here we
take exposures from all ages), to get $E(y)$ and then dividing:

\begin{equation}
f^\star(y) = \frac{B(y}{E(y)}
\end{equation}

We are interested to use thanatological fertility rates, $f^\star(y)$ in
models of population renewal,\footnote{This is probably premature, but helps to
understand the nature and provenance of model components.} rather than to
explicitly study the nature of these rates. Nonetheless it is best not to throw such rates into a model blindly, so we offer a schematic overview of some of their characteristics.
Figure~\ref{fig:FySpaghetti} represents the full variety in
female thanatological period fertility rates that can be found for all years
of data that overlap in the HFD and HMD.\footnote{There are as of this writing
1834 population-years of overlap between the HMD and HFD, including a wide
variety of fertility and mortality combinations.} Figure~\ref{fig:FxSpaghetti} gives ASFR for the same
populations and years as a more familiar reference, but note that both scales
are different. With no indication of particular populations or time series, one
already concludes that period thanatological fertility rates have a
characteristic shape and are not random, erratic or informationless: There is a
pattern to fertility rates by remaining years of life, and it varies over time
and between populations within some range of normality. Thanatological fertility
rates have a wider distribution than do chronological rates. Note that this
spaghetti plot includes both fertility booms and busts, as well as some
mortality crises (1918, WWII). Some patterns to note, but which we do not
separate graphically here, are that the left tail (a gauge of how orphan-prone a
population is) has tended to fall over time. All populations have shown a
rightward shift in both the mean and modal mother's thanatological age at birth
(TAB) over time. Several
populations now show modal TABs of over 60 years. Thanatological total fertility rates (TTFR)
track standard period total fertility rates (PTFR) rather closely, but tend to
be somewhat higher (not shown).

@@

In general, thanatological fertility rates will not be useful for purposes of
projection, unless the demographer believes that their empirical regularity is
somehow stronger than typical ASFR. This is a question we do not seek to answer
in the present treatment. There are alternative ways of defining
thanatological fertility rates, for instance by transforming the birth and
population vector from a stable population. As proven in appendix~\ref{app:B},
thanatological fertility rates derived from a stable population imply a certain
symmetry with stable populations from the chronological model. First let us
define the basic model.

\FloatBarrier
\section*{The thanatological renewal model}
For the renewal model that follows, we are most interested in using the kind
of fertility rates shown in Figure~\ref{fig:FySpaghetti}, and we refer to a
closed, unisex population based on female vital rates. Assuming constant vital rates, the
births for the present year are given by 
\begin{align}
B(t) =& \int _0^\infty P(y,t)f^\star(y) \dd y = \int _0^\infty P(a,t)f(a) \dd a
\quad\text{,} \intertext{where $a$ indexes chronological age, $y$ indexes
thanatological age, $P(a), P(y)$ are population counts, and $f^\star(y), f(a)$ are
exact specific fertility probabilities (rates), which here are best thought of
as the fertility rates of females born to females. The thanatological integral can
be broken down back in terms of chronological age:} =& \int_{y=0}^\infty
\int_{a=0}^\infty f^\star(y)\frac{P(a,t)d(a+y)}{l(a)} \dd a \dd y \quad\text{,}
\intertext{where $d(a)$ is the continuous lifetable death distribution with
radix of 1, so $d(a)/l(0)$. We can relate the present population to past births
with $P(a,t) = B(t-a)l(a)$:}
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t-a)d(a+y)\dd a \dd y
\quad\text{.} \intertext{Eventually strong ergodicity will assert itself, and $B(t)$ will be related to
$B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth:}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\quad\text{.}
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation:}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) d(a+y)e^{-ra}\dd a \dd y
\quad\text{.}
\end{align}
Now compare this to Lotka's chronological formulation
\begin{align}
\label{eq:lotka}
1 =& \int_{a=0}^\infty f(a)l(a)e^{-ra}\dd a \quad\text{,}
\intertext{which can be expressed as well as:}
  =& \int_{a=0}^\infty \int_{y=0}^\infty f(a)d(a+y)e^{-ra}\dd y \dd a
  \label{eq:lotka2}
\end{align}
and note that \eqref{eq:lotka2} and \eqref{eq:thanoren} are really quite
similar, since $\int _{y=0}^\infty d(a+y)\dd a = l(a)$. For intuition, notice that $l(a)$ is here split
up into pieces of $d(a)$, and imagine a 2D surface of these, where one axis is chronological age and the
other axis is thanatological age. For the chronological case \eqref{eq:lotka},
we multiply chronological age-specific fertility rates, $f(a)$, over the
chronological age margin, and for the thanatological case we multiply
$f^\star(y)$ over the thanatological age margin.

There are gaps in the above line of development, since the jump from
\eqref{eq:ergostep1} to \eqref{eq:ergostep2} (strong ergodicity) is unproven,
although it is rather intuitive, given the much greater density of connections
within the model; persons from nearly any thanatologcal age can produce offspring that can have any other thanatological age.\footnote{I say
\textit{nearly} because the very highest thanatological ages tend to be
composed of pre-menarchical females, so these can be thought of as structural
zeros. I am uncertain if it is a record, but it is certainly an extreme case
useful for illustration: Jeanne Calment, who lived to 122, had a daughter at age
22, which means she was about thanatological age 100
(\url{http://en.wikipedia.org/wiki/Jeanne_Calment}). If the lifetable used to
transform to thanatological closes out at age 110+, as is the case with HMD data, then fertility rates for thantological ages 101-110+ would be structural
zeros.} In
this sense, the smoothing mechanism at play as the population passes through
time must be much stronger than that for chronological age, at least in most
cases.  The mechanisms at play unfold in the same way as those so intuitively
described by \citet{arthur1982ergodic}, and said proof may apply here without
further modification. Later in this paper I note how strong ergodicity is
guaranteed given the properties of the thanatolgical projection matrix. A proof
of the uniqueness of the solution to \eqref{eq:thanoren} is given in
Appendix~\ref{app:A}. A proof that the chronological and thanatological renewal
models must yield the same $r$ if the starting population is already stable is
given in Appendix~\ref{app:B}. A fast-converging iterative method for finding
$r$ from equation~\eqref{eq:thanoren} is given in Appendix~\ref{app:C}

If the thanatolgical fertility rates do not derive from a stable
population, as will typicallybe the case, then the two age perspectives
will, unless by coincidence, yield divergent stable models. There is a strong
parallel here with the case of two-sex age-specific fertility rates. As in the
case of divergence between male and female single-sex models, there will
virtually always be divergence between single-sex models under chronological
versus thanatological age, unless the starting population is stable. Even though the modeled population stocks are in a way commensurable, the rates used here, $f(a)$ versus $f^\star(y)$ are calculated on the basis of differently distributed denominators $E(a)$ versus $E(y)$.

@@

Thanatological projection matrices may be manipulated using
standard matrix techniques applicable to the Leslie matrix. Where $\textbf{p}$
is a population vector classified by thanatological age, projection proceeds by
multiplying $\textbf{Y}$ from the left:
$\textbf{p}(t + 1) = \textbf{Y}\textbf{p}(t)$.

For rate schedules typical of human populations, thanatological projection
matrices are irreducible and primitive, and so will always have a
dominant positive real eigenvalue, the natural log of which is $r$. This
observation is sufficient to demonstrate the strong ergodic
theorem:\footnote{This step was taken for granted in jumping from
equation~\eqref{eq:ergostep1} to ~\eqref{eq:ergostep2}} i.e., given a fixed
matrix and enough time, the population will conform to some stable thanatological age distribution. The eigenvector corresponding to the dominant eigenvalue gives this stable thanatological age structure. The square
of the, or the pace of convergence to stability (the ratio of the \nth{1} to the \nth{2} eigenvalues).
\footnote{See \citet[p.86-87]{caswell2001matrix}.}

@@

\section{An iterative method to find the thanatological $r$}
\label{app:C}
\citet{coale1957new} proposed a fast-converging iterative approach to estimate
the intrinsic growth rate for the Lotka-Euler equation. For the thanatological
renewal model, a similar approach may be taken, with some slight
modifications to Coale's original.
The following steps can be followed to estimate $r$ from
Equation~\eqref{eq:thanoren}:
